% This must be in the first 5 lines to tell arXiv to use pdfLaTeX.
\pdfoutput=1

\documentclass[11pt]{article}

% Remove the "review" option to generate the final version.
\usepackage[review]{ACL2023}

\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{booktabs}
\usepackage{graphicx}

\title{Diagnose Before Simplifying: Operation-Aware Biomedical Text Simplification}

% TODO: fill in real author list / affiliations before submission
\author{
  Sruthilaya \and Rishabh \and Sophakotra \and Zihao \\
  Northeastern University \\
  \texttt{email@domain}
}

\begin{document}
\maketitle

\begin{abstract}
% TODO (team): 3-5 sentence summary. Draft below covers Sruthilaya's
% contribution; extend with classifier/generation results once available.
Biomedical text simplification systems typically apply a single, undifferentiated
simplification strategy to all text, regardless of what kind of complexity a given
span actually exhibits. We argue that classifying the \emph{operation} a span
requires -- Substitution, Explanation, or Generalization -- before generation
produces safer and more capable simplification than unguided end-to-end approaches.
We show that rule-based simplification tools exhibit a substantial coverage gap:
they cannot perform Explanation or Generalization operations at all, only
Substitution. We introduce a multi-signal complexity detection framework spanning
lexical, syntactic, numerical, and safety-critical dimensions, grounded in a
validated real-UMLS terminology lookup achieving complete jargon coverage on our
training set (100.0\%, vs. 86.8\% for a morphological heuristic baseline). We
further show that safety evaluation for simplification systems requires
warning-stratified sampling, not random sampling, to reliably detect rare but
safety-critical failures.
\end{abstract}

\section{Introduction}
% TODO (team): high-level framing, prior work context, contributions list.
% Suggested flow:
%  - Biomedical text is written for clinicians, not patients (cite PLABA motivation)
%  - Prior simplification work treats "simplify" as one operation; PLABA's own
%    taxonomy (Substitution/Explanation/Generalization/Omission/Exemplification)
%    shows this is wrong
%  - Our contribution: (1) operation-aware framing, (2) multi-signal complexity
%    detection incl. validated real-UMLS lookup, (3) stratified safety evaluation
%    methodology, (4) empirical coverage-gap quantification
Patient-facing biomedical text -- research abstracts, clinical notes, drug
labeling -- is frequently written at a reading and vocabulary level that assumes
clinical training the intended reader does not have. Automated simplification
systems aim to close this gap, but most treat simplification as a single,
undifferentiated transformation applied uniformly across a text. We argue this is
the wrong framing: different spans of biomedical text require fundamentally
different \emph{kinds} of simplification, and diagnosing which kind is needed
before generating output is itself the key missing step in prior work.

% TODO (team): insert contributions list, e.g.:
% Our contributions are: (1) ...; (2) ...; (3) ...; (4) ...

\section{The Problem}
% TODO (team): motivate why this is interesting/important beyond "simplification is nice"
% Suggested content: coverage gap statistic, safety stakes of getting it wrong
% (dropped contraindication vs. dropped adjective), PLABA operation distribution
% from Ondov et al. 2025 as evidence operations are not interchangeable.

\section{Main Idea}
% TODO (team): overall operation-aware architecture description belongs here
% (Sophakotra's classifier + generation routing). Sruthilaya's complexity
% detection framework and QuickUMLS integration below are the diagnostic
% front-end that motivates and feeds this routing decision.

\subsection{Complexity Detection Framework}
\label{sec:complexity-detection}

Before a biomedical sentence can be simplified, the system must first determine
\emph{what kind} of complexity it exhibits. We argue that biomedical text
complexity is not a single dimension but a composite of at least four distinct
signals, each requiring a different detection paradigm:

\begin{itemize}
    \item \textbf{Numerical expressions} (dosages, risk ratios, $p$-values,
    confidence intervals), detected via regular expressions. These patterns are
    finite and rule-governed, so no learned model is needed to achieve full
    recall on known formats.
    \item \textbf{Safety-critical warning language} (contraindications, adverse
    effects, risk alerts), detected via a hand-built lexicon. Because these
    phrases are clinically standardized, a lexicon gives full auditability and
    zero hallucination risk compared to a learned classifier.
    \item \textbf{Named biomedical entities} (diseases, chemicals), detected via
    a supervised NER model (SciSpaCy, BC5CDR-trained; \citealp{Neumann2019}).
    General-domain NER models miss domain-specific entities such as
    \emph{nephrotoxicity}; biomedical pretraining is not optional for this
    signal.
    \item \textbf{Structural complexity}, detected via dependency parse depth
    rather than raw sentence length. Length alone unfairly penalizes
    long-but-simple sentences (e.g., coordinated lists of facts), whereas parse
    depth captures genuine syntactic nesting that a vocabulary-only view misses
    entirely.
    \item \textbf{Medical jargon}, detected via UMLS concept matching
    (Section~\ref{sec:quickumls}). A term is treated as jargon if it exists in
    UMLS as a formal concept but is not part of common English vocabulary.
\end{itemize}

No single paradigm captures all five signals: regular expressions catch numbers
but miss entities; a lexicon catches warnings but misses syntactic structure; NER
catches entities but misses jargon that is not a named entity in the BC5CDR sense
(e.g., procedure names, qualifiers). We treat this as the central architectural
argument motivating a multi-detector design rather than a single end-to-end
classifier.

\subsection{QuickUMLS Integration}
\label{sec:quickumls}

Our jargon detector is built on the Unified Medical Language System (UMLS), a
meta-thesaurus unifying over 200 source vocabularies (SNOMED CT, MeSH, RxNorm,
and others) into a single concept space of several million entries. We use
QuickUMLS \citep{Soldaini2016}, which maps free text to UMLS concept identifiers
via approximate string matching, to perform this lookup at the sentence level.

\paragraph{Definition of jargon.} A term is flagged as jargon if it resolves to a
UMLS concept \emph{and} is not part of common English vocabulary, operationalized
via three filters layered on top of raw QuickUMLS output:

\begin{enumerate}
    \item \textbf{Semantic type restriction.} UMLS concepts are tagged with one
    or more of 127+ semantic types. We restrict matches to eight clinically
    relevant types (e.g., \textit{Disease or Syndrome}, \textit{Pharmacologic
    Substance}, \textit{Therapeutic or Preventive Procedure}), excluding generic
    administrative or qualifier concepts that are technically present in UMLS
    but are already plain English (e.g., \textit{primary}, \textit{performed}).
    \item \textbf{Similarity threshold.} We require a QuickUMLS similarity score
    of at least 0.85, filtering out loose or uncertain string matches.
    \item \textbf{Lexical frequency filter.} Semantic-type restriction alone is
    insufficient: some common English words coincidentally match real,
    clinically-relevant UMLS concepts. For example, \emph{control} matches a
    genuine pharmacologic-substance concept at similarity 1.0 -- the same
    semantic type needed to correctly catch real drug names such as
    \emph{sorafenib}. We therefore apply a word-frequency filter using Zipf
    frequency scores \citep{wordfreq}, requiring single-word matches to fall
    below a threshold of 3.8 to be treated as jargon. This threshold was
    validated against a held-out set of 16 known common and jargon terms,
    cleanly separating common words (\emph{control}: 5.40, \emph{determine}:
    4.63, \emph{disease}: 4.90) from genuine jargon (\emph{carcinoma}: 3.15,
    \emph{hyperglycemia}: 2.31, \emph{sorafenib}: 1.43). A preliminary attempt
    using a broad English wordlist (NLTK's \texttt{words} corpus) was rejected:
    it classified genuine jargon such as \emph{hyperglycemia} as ordinary
    English, since raw dictionary membership does not capture actual usage
    frequency.
\end{enumerate}

Additionally, since QuickUMLS returns every candidate concept considered for a
matched span rather than a single best match, we deduplicate by retaining only
the highest-similarity candidate per span.

\section{Details, Evaluation, and Results}
% TODO (team): classifier results (Sophakotra), full-corpus eval (Zihao),
% readability-vs-safety scatter plot (Rishabh) go here alongside the
% subsections below.

\subsection{Safety Evaluation Methodology}
\label{sec:safety-eval}

We measure whether a simplification system preserves safety-critical content
(contraindications, adverse effect warnings) using a warning-preservation metric
with paraphrase matching, rather than surface $n$-gram overlap metrics such as
ROUGE. ROUGE cannot distinguish between a system that preserves a warning and one
that silently drops it, provided the rest of the sentence is scored favorably;
our metric targets exactly this failure mode.

A central methodological finding from our evaluation is that \textbf{random
sampling is statistically underpowered to detect rare safety failures}.
Warning-bearing sentences occur at a base rate of approximately 33\% in our data;
a random sample of 50 sentences showed all baseline systems achieving perfect
warning preservation, a misleadingly optimistic result. Only a
\emph{stratified} stress test -- restricted to warning-bearing sentences
specifically -- revealed that unguided LLM-based simplification is the only
baseline that drops safety-critical content under evaluation. We treat
stratified, targeted evaluation as a required methodological component for
safety-critical NLP systems generally, not specific to this dataset.

\subsection{Complexity Coverage Results}
\label{sec:complexity-results}

Table~\ref{tab:complexity-coverage} reports detector flag rates across our
verified training split ($n=530$ abstracts\footnote{We note a discrepancy
between this figure and an earlier internal estimate of $n=635$; the current
figure was independently re-derived via two methods (direct exclusion of
validation/test PMIDs from the source data, and the shipped train/validation/test
split file) which agree with each other. We report the verified, reproducible
number here.}), comparing our real UMLS-based jargon detector against a
morphological (suffix-pattern) fallback baseline.

\begin{table}[t]
\centering
\small
\begin{tabular}{lrr}
\toprule
\textbf{Detector} & \textbf{Morph.} & \textbf{UMLS} \\
\midrule
NER              & --    & 98.7\% \\
Warning cues      & --    & 32.1\% \\
Syntactic depth   & 94.0\% & 94.0\% \\
Numerical exprs.  & --    & 60.2\% \\
Jargon (UMLS)     & 86.8\% & \textbf{100.0\%} \\
\midrule
3+ detectors firing & 90.8\% & 98.1\% \\
\bottomrule
\end{tabular}
\caption{Detector flag rates by abstract ($n=530$). ``Morph.'' denotes a
suffix/pattern heuristic fallback used when a full UMLS installation is
unavailable; ``UMLS'' denotes the full QuickUMLS pipeline described in
Section~\ref{sec:quickumls}.}
\label{tab:complexity-coverage}
\end{table}

The real UMLS pipeline achieves complete (100.0\%) jargon coverage across the
training set, compared to 86.8\% under the morphological fallback -- a direct,
quantified demonstration of the coverage gap between rule-based jargon detection
and true terminology-base lookup. We validated this result in two stages: first
against a small set of targeted test sentences containing known jargon (e.g.,
\emph{hepatocellular carcinoma}, \emph{sorafenib}, \emph{nephrotoxicity}), and
second via manual review of a random sample of ten training abstracts, confirming
that flagged terms were genuine clinical terminology (e.g., \emph{implantable
cardioverter-defibrillator}, \emph{hyporeninemic hypoaldosteronism},
\emph{gabapentinoids}) rather than residual false positives.

Consistent with our preliminary findings, the majority of abstracts exhibit
multiple simultaneous complexity signals: 98.1\% of abstracts trigger three or
more of our five detectors under the full UMLS pipeline, reinforcing that
biomedical text complexity is multi-dimensional and that a single-signal
detector would systematically under-detect the true complexity profile of this
domain.

\section{Related Work}
% TODO (team): condense literature review to fit page budget. Suggested set:
\begin{itemize}
    \item \citet{Ondov2025} -- PLABA track creators' own reflection; source of
    the correct 3-operation taxonomy scope and an external classifier benchmark.
    \item \citet{Attal2023} -- original PLABA dataset and operation taxonomy.
    \item Alva-Manchego et al. 2025 -- TSAR shared task findings; no prior
    system classifies complexity type before generating.
    \item \citet{Neumann2019} -- SciSpaCy, justifies biomedical NER pretraining.
    \item Miyata et al. 2025 -- EhiMeNLP; shows generate-then-rerank gains come
    from diversity/selection, not operation understanding, motivating our
    diagnostic framing.
\end{itemize}

\section{Conclusion and Future Work}
% TODO (team): summarize contributions, point to classifier upgrade path
% (TF-IDF+LR -> BERT -> BioBERT), entity/numerical preservation metrics,
% human evaluation.

\section*{Limitations}
Our jargon detection pipeline, while substantially more complete than a
morphological baseline, retains residual imprecision at the phrase level:
multi-word spans (e.g., \emph{urine output}) are not subject to our single-word
frequency filter and can be flagged despite being largely comprehensible to a lay
reader. We view this as an acceptable trade-off given diminishing returns on
further manual filtering, but note it explicitly rather than presenting our
jargon coverage figure as free of false positives. Our warning-preservation
metric's paraphrase map encodes annotator assumptions about semantic equivalence
and would benefit from a learned semantic-similarity replacement (e.g.,
BioBERT-based scoring) in future work.
% TODO (team): add classifier/generation-specific limitations (macro-F1 ceiling,
% English-only scope, etc.)

\section*{Ethics Statement}
% TODO (team): broader impact of automated biomedical simplification --
% risk of dropped safety information, need for human oversight before
% clinical deployment, etc.

\bibliography{anthology,custom}
\bibliographystyle{acl_natbib}

\end{document}